---
title: "Syntopy in North American Passerines"
subtitle: "Spatial Analysis of BBS Data with R"
date: last-modified
categories: [Syntopy, Spatial Data, Birds, R, BBS]
author: "Matěj Tvarůžka"
format: html
---

### Materials and Methods

The study will use data from the **North American Breeding Bird Survey (BBS)**, a long-term, standardised monitoring programme that records bird assemblages across approximately 2,900 routes annually. Each route includes 50 point counts spaced at 0.5-mile intervals. I will analyse patterns of syntopy at two spatial resolutions:

1.  **Point scale (individual stops) and**

2.  **Transect scale** (combined stops along each route).

To infer drivers of syntopy, I will replicate the analytical framework of Remeš & Harmáčková (2023), which uses cooccurrence analyses and species trait data to identify ecological predictors of coexistence. A novel feature of this thesis will be analysing syntopy at two spatial resolutions (see above).

*Remeš, V. and Harmáčková, L. (2023), Resource use divergence facilitates the evolution of secondary syntopy in a continental radiation of songbirds (Meliphagoidea): insights from unbiased co-occurrence analyses. Ecography, 2023: e06268. https://doi.org/10.1111/ecog.06268*

# Sister pairs extraction

In this code I will select sister species pairs of North American passerines for the subsequent analysis of syntopy.

```{r}
#| label: Libraries
#| include: false
require(diverge)
require(clootl)
require(tidyverse)
require(ape)
```

Firstly we are going to select the phylogeny of wood-warblers to try the workflow. We can get the acces to eBird taxonomy files using `taxonomyGet()`, than we select the family of **wood-warblers** (*Parulidae*).

```{r}
#| include: false
#| label: Taxonomy
taxonomy <- taxonomyGet(taxonomy_year = 2021)
warblers <- taxonomy |> 
  filter(FAMILY == "Parulidae (New World Warblers)")
head(warblers)
```

We need to select the species names vector for the tree extraction:

```{r}
#| echo: true
#| label: Species names vector
warb_names_vec <- warblers |> 
  select(SCI_NAME) |> 
  pull(SCI_NAME)
glimpse(warb_names_vec)
```

We can extract the tree using `extractTree()` function from `clootl` package:

```{r}
#| label: Tree extraction
tree <- extractTree(species = warb_names_vec, taxonomy_year = 2021)
tree
```

We can plot it using `ape` package.

```{r}
plot(tree, type = "fan", cex = 0.5, tip.color = "darkblue")
```

Now it's time to extract species pairs:

```{r}
sister_pairs <- extract_sisters(tree)
head(sister_pairs)
```
