---
title: "Sister pairs selection for syntopy analysis"
author: "Matěj Tvarůžka"
format: html
---

# Sister pairs extraction

In this code I will select sister species pairs of North American passerines for the subsequent analysis of syntopy.

```{r}
#| label: Libraries
#| include: false
require(diverge)
require(clootl)
require(tidyverse)
require(ape)
```

Firstly we are going to select the phylogeny of wood-warblers to try the workflow. We can get the acces to eBird taxonomy files using `taxonomyGet()`, than we select the family of **wood-warblers** (*Parulidae*).

```{r}
#| include: false
#| label: Taxonomy
taxonomy <- taxonomyGet(taxonomy_year = 2021)
warblers <- taxonomy |> 
  filter(FAMILY == "Parulidae (New World Warblers)")
head(warblers)
```

We need to select the species names vector for the tree extraction:

```{r}
#| echo: true
#| label: Species names vector
warb_names_vec <- warblers |> 
  select(SCI_NAME) |> 
  pull(SCI_NAME)
glimpse(warb_names_vec)
```

We can extract the tree using `extractTree()` function from `clootl` package:

```{r}
#| label: Tree extraction
tree <- extractTree(species = warb_names_vec, taxonomy_year = 2021)
tree
```

Now it's time to extract species pairs:

```{r}
sister_pairs <- extract_sisters(tree)
head(sister_pairs)
```

