---
title: "Selection of sympatric sister species"
author: "Matěj Tvarůžka"
format:
  html:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Table of contents
toc-depth: 4
toc-location: left
execute: 
  eval: true
  echo: true
output:
  html_document:
  output-file: cooccurrence_data_prep.html
---

1. Sister pairs extraction from phylogeny -----------------------------------

Libraries

```{r}
# Lines 4-7
require(diverge)
require(clootl)
require(tidyverse)
library(sf)
```

Selecting taxonomy from eBird using clootl package
Selecting the vector of scientific names

```{r}
# Lines 11
taxonomy <- taxonomyGet(taxonomy_year = 2023)
```

Selecting passerine family

```{r}
# Lines 13-16
sci_names_vec <- taxonomy |> 
  filter(ORDER1 == "Passeriformes") |> 
  select(SCI_NAME) |> 
  pull(SCI_NAME)
```

Selecting the vector of species codes

```{r}
# Lines 19-22
codes_vec <- taxonomy |> 
  filter(ORDER1 == "Passeriformes") |> 
  select(SPECIES_CODE) |> 
  pull(SPECIES_CODE)
```

This replaces any digit (0-9) with an empty string

```{r}
# Lines 25
clean_codes <- gsub("[0-9]", "", codes_vec)
```

Extract the tree

```{r}
# Lines 28-29
tree_sci_names <- extractTree(species = sci_names_vec, taxonomy_year = 2023)
is.ultrametric(tree_sci_names) # False
```

Extract sister species pairs using extract_sisters() function from diverge package

```{r}
# Lines 31
sci_names_sister <- extract_sisters(tree_sci_names)
```

Selecting with species codes doesn't work
tree_codes <- extractTree(species = clean_codes, taxonomy_year = 2023)
codes_sister <- extract_sisters(tree_codes)

Inspect the results

```{r}
# Lines 38
head(sci_names_sister)
```

2. Identification of available species data - ebirdst & BBS ----------------------
With the sister pairs of passerines extracted, we can look which of those are presented in eBird S&T and BBS data
==== eBird S&T

```{r}
# Lines 43
library(ebirdst)
```

Get the list of species that actually have eBird S&T models

```{r}
# Lines 46-48
available_ebird_models <- ebirdst_runs |> 
  select(species_code, scientific_name) |> 
  mutate(sci_name_underscore = gsub(" ", "_", scientific_name))
```

Create a lookup vector for quick checking

```{r}
# Lines 51-59
ebird_species_list <- available_ebird_models$sci_name_underscore
sister_pairs_ebirdst <- sci_names_sister |> 
  # Check if Species 1 is in eBird S&T
  filter(sp1 %in% ebird_species_list) |> 
  # Check if Species 2 is in eBird S&T
  filter(sp2 %in% ebird_species_list)
message("Final N of sister pairs with eBird models: ", nrow(sister_pairs_ebirdst))
```

So we have about 262 sister pairs of passerines from ebirdst

==== BBS
Selecting species list from 2019 released dataset → downloaded with bbsAssistent() package

```{r}
# Lines 64-78
bbs_species_list <- bbs$species_list |> 
  select(Scientific_Name) |> 
  mutate(sci_name_underscore = gsub(" ", "_", Scientific_Name))
bbs_species_list_vec <- bbs_species_list$sci_name_underscore
sister_pairs_bbs <- sister_pairs_ebirdst |> 
  # Check if Species 1 is in BBS
  filter(sp1 %in% bbs_species_list_vec) |> 
  # Check if Species 2 is in BBS
  filter(sp2 %in% bbs_species_list_vec)
check <- unique(c(sister_pairs_bbs$sp1, sister_pairs_bbs$sp2))
summary(check) # each column has a unique set of species
message("Final N of sister pairs: ", nrow(sister_pairs_bbs))
```

The max count of sister pairs for BBS and ebird S&T data overlap is 54

3. Selecting ranges using ebirdst ----------------------------

Create a lookup table using the internal ebirdst taxonomy

```{r}
# Lines 84-87
spp_names_clean <- gsub("_", " ", sister_pairs_bbs)
ebird_lookup <- ebirdst_runs |>  
  filter(scientific_name %in% spp_names_clean) |> 
  select(species_code, scientific_name)
```

Extract codes vector

```{r}
# Lines 90
codes_to_download <- ebird_lookup$species_code
```

An access key is required to download eBird Status and Trends data
set_ebirdst_access_key("56fnsc0aep49")

Download the data (ranges only to save space/time)
for (code in codes_to_download) {
ebirdst_download_status(
species = code,
path = ebirdst_data_dir(),
download_abundance = FALSE,
download_occurrence = FALSE,
download_count = FALSE,
download_ranges = TRUE,
download_regional = FALSE,
download_pis = FALSE,
download_ppms = FALSE,
download_all = FALSE,
pattern = NULL,
dry_run = FALSE,
force = FALSE,
show_progress = TRUE
)
}

Load the breeding ranges (BBS data represents the breeding season)

```{r}
# Lines 116-125
ranges <- lapply(codes_to_download, function(code) {
  load_ranges(species = code,
              resolution = c("9km", "27km"),
              smoothed = TRUE,
              path = ebirdst_data_dir()
  )
})
class(ranges)
ranges[[1]]
```

4. Range overlap calculation -----------------------------------------------

```{r}
# Lines 129-157
names(ranges) <- codes_to_download
calculate_overlap <- function(code1, code2, range_list) {
  # 1. Extract polygons
  s1 <- range_list[[code1]]
  s2 <- range_list[[code2]]
  # 2. Filter for Breeding OR Resident
  s1_b <- s1[s1$season %in% c("breeding", "resident"), ]
  s2_b <- s2[s2$season %in% c("breeding", "resident"), ]
  # 3. If either is missing a range, return 0
  if (nrow(s1_b) == 0 | nrow(s2_b) == 0) return(0)
  # 4. Calculate areas (st_area returns units, so we convert to numeric)
  area1 <- as.numeric(st_area(s1_b))
  area2 <- as.numeric(st_area(s2_b))
  # 5. Find intersection
  # st_make_valid handles geometry errors often found in complex range maps
  inter <- st_intersection(st_make_valid(s1_b), st_make_valid(s2_b))
  if (nrow(inter) == 0) {
    return(0)
  } else {
    area_inter <- sum(as.numeric(st_area(inter)))
    # Index: Intersection / smaller of the two ranges
    return(area_inter / min(area1, area2))
  }
}
```

Convert names to match scientific_name in ebird_lookup

```{r}
# Lines 160-166
sister_pairs <- sister_pairs_bbs |>
  mutate(sp1_clean = gsub("_", " ", sp1),
         sp2_clean = gsub("_", " ", sp2)) |>
  left_join(ebird_lookup, by = c("sp1_clean" = "scientific_name")) |>
  rename(code1 = species_code) |>
  left_join(ebird_lookup, by = c("sp2_clean" = "scientific_name")) |>
  rename(code2 = species_code)
```

Run the calculation across all 54 pairs

```{r}
# Lines 169-187
sister_pairs$range_overlap <- sapply(1:nrow(sister_pairs), function(i) {
  calculate_overlap(
    sister_pairs$code1[i], 
    sister_pairs$code2[i], 
    ranges
  )
})
### It gives this warning:  attribute variables are assumed to be spatially constant throughout all geometries
ggplot(sister_pairs, aes(x = range_overlap)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white", boundary = 0) +
  theme_minimal() +
  labs(
    title = "Distribution of Range Overlap in 54 Passerine Sister Pairs",
    x = "Symmetric Overlap Index (0 = Allopatric, 1 = Completely Overlapping)",
    y = "Number of Pairs"
  ) +
  scale_x_continuous(breaks = seq(0, 1, 0.1))
sum(sister_pairs$range_overlap > 0, na.rm = TRUE)
```

45 pairs have some degree of range overlap

```{r}
# Lines 189-190
sister_pairs$sympatry <- sister_pairs$range_overlap >= 0.05
sum(sister_pairs$sympatry == T)
```

34 pairs are sympatric for 5 % treshold

5. Inspecting N of BBS routes in sympatric zone -------------------------

Convert BBS routes to a spatial object (sf)

```{r}
# Lines 196-217
bbs_routes_sf <- st_as_sf(bbs$routes, 
                          coords = c("Longitude", "Latitude"), 
                          crs = 4326)
count_routes_sympatry <- function(code1, code2, range_list, routes_sf) {
  # Function to count routes in sympatric range
  # Get breeding ranges
  r1 <- range_list[[code1]][range_list[[code1]]$season %in% c("breeding", "resident"), ]
  r2 <- range_list[[code2]][range_list[[code2]]$season %in% c("breeding", "resident"), ]
  if(nrow(r1) == 0 | nrow(r2) == 0) return(0)
  # Find the intersection polygon
  overlap_poly <- st_intersection(st_make_valid(r1), st_make_valid(r2))
  if(nrow(overlap_poly) == 0) return(0)
  # Count routes that fall inside this polygon
  routes_inside <- st_intersects(routes_sf, overlap_poly, sparse = FALSE)
  return(sum(routes_inside))
}
```

lookup for AOU codes using sister_pairs names

```{r}
# Lines 220-222
aou_lookup <- bbs$species_list |> 
  mutate(sci_name_underscore = gsub(" ", "_", Scientific_Name)) |> 
  select(AOU, sci_name_underscore)
```

Add AOU codes to sister_pairs

```{r}
# Lines 225-252
sister_pairs <- sister_pairs |> 
  left_join(aou_lookup, by = c("sp1" = "sci_name_underscore")) |> 
  rename(aou1 = AOU) |> 
  left_join(aou_lookup, by = c("sp2" = "sci_name_underscore")) |> 
  rename(aou2 = AOU) |> 
  select(-sp1_clean, sp2_clean)
names(sister_pairs)
count_shared_routes <- function(a1, a2, obs_df) {
  # Function to count "Active" shared routes
  # Find routes where Sp1 was ever seen
  r_sp1 <- obs_df$RouteDataID[obs_df$AOU == a1]
  # Find routes where Sp2 was ever seen
  r_sp2 <- obs_df$RouteDataID[obs_df$AOU == a2]
  # Intersection of Route IDs
  shared_routes <- intersect(r_sp1, r_sp2)
  return(length(unique(shared_routes)))
}
d <- sister_pairs |>
  filter(sympatry == T) |> 
  mutate(
    sympatry_routes = map2_dbl(code1, code2, ~count_routes_sympatry(.x, .y, ranges, bbs_routes_sf)),
    bbs_shared_routes = map2_dbl(aou1, aou2, ~count_shared_routes(.x, .y, bbs$observations))
  )
head(d, 5)
```

